---
title: "NCDF Report for Kisii, Nyamira and Kajiado"
author: "Everlyne N. Angwenyi"
date: "2026-02-05"
mainfont: Times New Roman
format:
    pdf:
      documentclass: report
      geometry:
        - margin=1in
      fontsize: 12pt
      linestretch: 2
      include-in-header:
        text: |
          \usepackage{etoolbox}
          \AtBeginEnvironment{verbatim}{\linespread{1}\selectfont\small}
          \AtBeginEnvironment{Highlighting}{\linespread{1}\selectfont\small}
      include-before-body: 
        text: |
          \begin{center}
          \text{SDS6/49662/2025}
          
          \vspace{0.5em}
          
          \text{Department of Mathematics}
          
          \text{Faculty of Science and Technology}
          
          \text{Statistical Computing - SDS 6103}
          \end{center}
          \vspace{2em}
editor: visual
bibliography: NCDF_references.bib
link-citations: true
---

\newpage

## Introduction

The National Government Constituencies Development Fund (NG‑CDF) is a key mechanism for financing grassroots development projects in Kenya. This study focuses on allocations in Kisii, Nyamira, and Kajiado counties between 2014 and 2026. The aim is to analyze allocation trends, compare constituency performance, and assess growth rates.

A Shiny dashboard was developed to provide stakeholders with interactive visualizations of allocations. For reporting purposes, this Quarto document integrates static plots and tables that summarize the findings. While allocations are clearly tracked, the analysis also highlights the gap in linking financial flows to actual projects implemented, which is essential for accountability and impact assessment.

## Methodology

The analysis was conducted using R and several supporting packages:

**tidyverse** for data manipulation and wrangling.

**scales** for formatting numeric outputs.

**ggplot2** for data visualization.

**knitr** for static tables.

```{r}
# Load required packages
library(tidyverse)   # includes dplyr, ggplot2, tidyr, readr, etc.
library(scales)
library(knitr)
```

**Steps** **Step 1: Load Data** Load the dataset containing NCDF allocations.

```{r}
NCDF <- read.csv("NEW_NGCDF_DATA_28-JAN_2026.csv", sep = ',', header = TRUE)
NCDF$County <- trimws(NCDF$County)
```

**Step 2: Filter Counties of Interest** Restricted the dataset to Kisii, Nyamira, and Kajiado.

```{r}
NCDF_County <- dplyr::filter(
  NCDF,
  County %in% c("Kisii", "Kajiado", "Nyamira"),
  County != 0
)
head(NCDF_County);
```

**Step 3: Reshape Data** Pivot yearly allocations into long format for easier analysis.

```{r}
NCDF_long <- NCDF_County %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "Year",
               values_to = "Allocation") %>%
  mutate(Year = as.numeric(sub("X", "", Year))) %>%
  filter(!is.na(Year))
head(NCDF_long);
```

**Step 4: Summarize Allocations** Computes county‑level totals per year.

```{r}
County_Summary <- NCDF_long %>%
  group_by(County, Year) %>%
  summarise(Total_Allocation = sum(Allocation, na.rm = TRUE), .groups = "drop")
head(County_Summary);
```

**Step 5: Visualization** Generates plots to show trends and comparisons.

```{r}
ggplot(County_Summary, aes(x = Year, y = Total_Allocation, color = County)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "NCDF Allocations (2014–2026)",
       x = "Year", y = "Total Allocation (KES)") +
  theme_minimal()
```

## Results and Analysis

**County-Level Trends**

```{r}
ggplot(County_Summary, aes(x = Year, y = Total_Allocation, color = County)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "NCDF Allocations (2014–2026)",
       x = "Year", y = "Total Allocation (KES)") +
  theme_minimal()
```

This line plot shows how allocations have evolved over time. Kisii and Nyamira display steady upward trends, while Kajiado’s allocations fluctuate more, suggesting variability in funding priorities.

**County Comparison (Total Allocations)**

```{r}
df <- County_Summary %>%
  filter(Year >= 2014, Year <= 2026) %>%
  group_by(County) %>%
  summarise(Total_Allocation = sum(Total_Allocation, na.rm = TRUE))

ggplot(df, aes(x = reorder(County, -Total_Allocation),
               y = Total_Allocation/1e6,
               fill = County)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Total_Allocation/1e6,1)," M")),
            vjust = -0.5, size = 3) +
  labs(title = "Comparison of NCDF Allocations (2014–2026)",
       y = "Allocation (Million KES)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

This bar chart compares total allocations across the three counties. It highlights which county received the highest cumulative funding over the study period.

**Cross-County Yearly Allocations**

```{r}
df <- NCDF_long %>%
  filter(Year >= 2014, Year <= 2026,
         County %in% c("Kisii", "Nyamira", "Kajiado"))

ggplot(df, aes(x = factor(Year),
               y = Allocation/1e6,
               fill = Constituency)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ County, scales = "free_y") +
  labs(title = "Yearly Allocations by Constituency (Million KES)",
       y = "Allocation (Million KES)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This faceted chart shows yearly allocations broken down by constituency within each county. It reveals which constituencies dominate funding in a given year and highlights intra‑county disparities.

**Side-by-Side County Comparison**

```{r}
ggplot(County_Summary, aes(x = factor(Year),
                           y = Total_Allocation/1e6,
                           fill = County)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "NCDF Allocations by County (2014–2026)",
       y = "Allocation (Million KES)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This grouped bar chart compares counties year by year. It makes clear whether one county consistently outpaces the others or whether funding shifts across years.

**Growth Rate Comparison**

```{r}
growth_df <- County_Summary %>%
  group_by(County) %>%
  summarise(GrowthRate = (last(Total_Allocation) - first(Total_Allocation)) /
                          first(Total_Allocation) * 100)

ggplot(growth_df, aes(x = reorder(County, GrowthRate),
                      y = GrowthRate,
                      fill = County)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(GrowthRate,1), "%")),
            vjust = -0.5) +
  labs(title = "Growth Rate of Allocations (2014–2026)",
       y = "Growth Rate (%)") +
  theme_minimal()
```

This chart shows the percentage growth in allocations from 2014 to 2026. It highlights which county experienced the fastest growth in funding, offering insight into changing budget priorities.

**Top Constituencies per Year**

```{r}
library(knitr)

top_constituencies <- NCDF_long %>%
  filter(County %in% c("Kisii", "Nyamira", "Kajiado"),
         Year >= 2014, Year <= 2026) %>%
  group_by(County, Year) %>%
  slice_max(order_by = Allocation, n = 1, with_ties = FALSE) %>%
  select(Year, County, Constituency, Allocation)

top_constituencies %>%
  kable(caption = "Top Constituency Allocations per County per Year (2014–2026)")
```

This table identifies the top‑funded constituency in each county for every year. It shows which constituencies consistently attract the highest allocations, and whether dominance shifts over time. Comparing across counties also reveals if one constituency stands out nationally among the three.

## Findings

The analysis of NCDF allocations across Kisii, Nyamira, and Kajiado between 2014 and 2026 reveals several key insights:

**County-level trends** - Kisii and Nyamira show steady increases in allocations, while Kajiado exhibits more variability, reflecting shifting priorities or budget constraints.

**Constituency comparisons** - Certain constituencies consistently receive higher allocations, suggesting prioritization or larger project scopes.

**Growth rates** - Growth rates vary widely, with some constituencies showing strong upward trends while others remain stagnant.

**Top allocations** - Each year, one constituency emerges as the top recipient in each county. Comparing across counties shows that Kisii constituencies often dominate, though Nyamira and Kajiado occasionally surpass them in specific years.

## Identified Gaps

Despite the dashboard’s strength in tracking financial allocations, several limitations remain:

**Project-level data is missing** — the dashboard does not show what projects were funded (schools, health centers, water points, roads).

**Completion status is not captured** — stakeholders cannot assess whether projects were completed, ongoing, or stalled.

**Impact assessment is absent** — the quality, sustainability, and community impact of funded projects are not evaluated. Accountability metrics such as planned vs. actual outputs, unit costs, and delays are not integrated.

## Recommendations

To strengthen accountability and transparency, the following enhancements are proposed:

**Integrate project**-level data - Extend the dataset to include project descriptions, categories, start/end dates, and completion status. Link allocations to specific outputs (e.g., “KES 20M → 3 classrooms built in Kisii South”).

**Web scraping for project information** -Use tools like rvest in R to collect project lists from NCDF Board, county portals, and parliamentary reports.

**Dashboard enhancements** - Add a Projects tab showing allocations linked to outputs.Use interactive maps to show project locations. Include completion status indicators (e.g., green = completed, red = stalled). Accountability metrics - Compare planned vs. actual projects.Track unit costs (e.g., cost per classroom).Highlight delays or discrepancies between allocations and outcomes.

**Stakeholder engagement** - Provide downloadable reports for community monitoring. Enable citizen feedback loops to validate project completion.

## Conclusion

The NCDF dashboard for Kisii, Nyamira, and Kajiado counties provides a strong foundation for financial transparency. The findings show clear allocation trends, county comparisons, and constituency‑level disparities. However, allocations alone do not capture the development outcomes.

To achieve accountability, the dashboard must evolve to include project‑level data. By integrating project descriptions, completion status, and impact metrics, stakeholders will be able to compare financial inputs with tangible outputs. This transformation will ensure that NCDF allocations are not only tracked as numbers but also evaluated in terms of their real impact on communities like empowering citizens, policymakers, and researchers to monitor progress and demand accountability.
