---
title: "NCDF Report for Kisii, Nyamira and Kajiado"
author: "Everlyne N. Angwenyi"
date: "2025-11-15"
mainfont: Times New Roman
format:
    pdf:
      documentclass: article
      geometry:
        - margin=1in
      fontsize: 12pt
      linestretch: 2
      include-in-header:
        text: |
          \usepackage{etoolbox}
          \AtBeginEnvironment{verbatim}{\linespread{1}\selectfont\small}
          \AtBeginEnvironment{Highlighting}{\linespread{1}\selectfont\small}
      include-before-body: 
        text: |
          \begin{center}
          \text{SDS6/49662/2025}
          
          \vspace{0.5em}
          
          \text{Department of Mathematics}
          
          \text{Faculty of Science and Technology}
          \end{center}
          \vspace{2em}
editor: visual
bibliography: example_references.bib
link-citations: true
---

\newpage
\tableofcontents
\newpage


# Abstract
This report presents an analysis of NG-CDF allocations in Kisii, Nyamira, and Kajiado counties between 2014 and 2026. Using R for data cleaning, transformation, and visualization, the study highlights county-level and constituency-level trends, compares performance, and evaluates growth rates. Interactive Shiny app outputs are integrated to complement static plots.

# Introduction
The NG-CDF is a key funding mechanism for grassroots development in Kenya. This assignment demonstrates practical skills in data wrangling, visualization, and interactive reporting by analyzing allocations in three counties of interest.

## Data Preparation
```{r}
library(tidyverse)
library(dplyr)

# Load Dataset
NCDF <- read.csv("NEW_NGCDF_DATA_28-JAN_2026.csv", sep = ',', header = TRUE)
NCDF$County <- trimws(NCDF$County)

# Filter for Counties of Interest
NCDF_County <- NCDF %>%
  filter(County %in% c("Kisii", "Kajiado", "Nyamira"), County != 0)

# Pivot Years into Long Format
NCDF_long <- NCDF_County %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "Year",
               values_to = "Allocation") %>%
  mutate(Year = as.numeric(sub("X", "", Year))) %>%
  filter(!is.na(Year))
```
#County-Level Trends
```{r}
County_Summary <- NCDF_long %>%
  group_by(County, Year) %>%
  summarise(Total_Allocation = sum(Allocation, na.rm = TRUE), .groups = "drop")

ggplot(County_Summary, aes(x = Year, y = Total_Allocation, color = County)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "NCDF Allocations (2014–2026)",
       x = "Year", y = "Total Allocation (KES)") +
  theme_minimal()
```
\newpage

# References
